---
title: "Penguin Data Analysis Portfolio"
author: "Isaac Wainaina"
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    code-fold: true
    page-layout: full
    highlight-style: github
    fig-cap-location: bottom
---

## Dataset overview

This portfolio explores the **Palmer Penguins** dataset to understand how penguin species differ in morphology and habitat. The goal is to practice data visualization, summary statistics, and exploratory data analysis using R and Quarto.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(palmerpenguins)
library(ggthemes)
library(data.table)
library(knitr)
```

## Descriptive statistics

### Summary Statistics of Key Measurements

@fig-summary shows summary statistics of key penguin measurements

```{r}
#| label: fig-summary
#| fig-asp: 0.618
#| fig-alt: Summary statistics of key penguin measurements

penguins |>
  select(
    "Bill Length (mm)" = bill_length_mm,
    "Bill Depth (mm)" = bill_depth_mm,
    "Flipper Length (mm)" = flipper_length_mm,
    "Body Mass (g)" = body_mass_g
  ) |>
  summary() |>
  matrix(ncol = 4, byrow = FALSE,
         dimnames = list(
           c("Min.", "1st Qu.", "Median", "Mean", "3rd Qu.", "Max.", "NA's"),
           c("Bill Length (mm)", "Bill Depth (mm)", "Flipper Length (mm)", "Body Mass (g)")
         )) |>
  kable(caption = "Summary statistics of key penguin measurements", align = "r")

```

The penguins in the dataset show distinct variation in body morphology:

-   **Bill length** ranges from around 32 mm to 60 mm, with an average of about 44 mm.

-   **Bill depth** averages approximately 17 mm, ranging from about 13 mm to 21 mm.

-   **Flipper length** spans from 172 mm to 231 mm, with a mean close to 201 mm.

-   **Body mass** varies widely, from 2700 g to 6300 g, averaging around 4200 g.

### Count by Species

@fig-spec-count shows the number of penguins by species

```{r}
#| label: fig-spec-count
#| fig-asp: 0.618
#| fig-alt: Number of Penguins by Species

table(penguins$species) |>
  as.data.frame() |>
  kable(
    col.names = c("Species", "Count"),
    caption = "Number of Penguins by Species"
  )

```

### Count by Island

@fig-isl-count shows number of penguins by island

```{r}
#| label: fig-isl-count
#| fig-asp: 0.618
#| fig-alt: Number of Penguins by Island

table(penguins$island) |>
   as.data.frame() |>
  kable(
    col.names = c("Island", "Count"),
    caption = "Number of Penguins by Island"
  )

```

### Count by Sex

@fig-sex-count shows number of penguins by sex

```{r}
#| label: fig-sex-count
#| fig-asp: 0.618
#| fig-alt: Number of Penguins by Sex

table(penguins$sex) |>
   as.data.frame() |>
  kable(
    col.names = c("Sex", "Count"),
    caption = "Number of Penguins by Sex"
  )
```

### Count by Species and Sex

@fig-sex-species-count number of Penguins by species and sex

```{r}
#| label: fig-sex-species-count
#| fig-asp: 0.618
#| fig-alt: Number of Penguins by Species and Sex

table(penguins$species, penguins$sex) |>
  as.data.frame() |>
  filter(!is.na(Var2)) |>
  pivot_wider(
    names_from = Var2,
    values_from = Freq,
    values_fill = 0
  ) |>
  rename(Species = Var1) |>
  kable(
    caption = "Penguin Count by Species and Sex (excluding unknown sex)"
  )
```

### Count by Sex and Island

@fig-isl-sex-count number of penguins by sex and island

```{r}
#| label: fig-isl-sex-count
#| fig-asp: 0.618
#| fig-alt: Number of Penguins by Sex and Island

table(penguins$sex, penguins$island) |>
  as.data.frame() |>
  pivot_wider(
    names_from = Var2,
    values_from = Freq,
    values_fill = 0
  ) |>
  rename(Sex = Var1) |>
  kable(
    caption = "Penguin Count by Sex and Island"
  )
```

### Count by Species and Island

@fig-isl-species-count number of penguins by species and island

```{r}
#| label: fig-isl-species-count
#| fig-asp: 0.618
#| fig-alt: Number of Penguins by Species and Island

table(penguins$species, penguins$island) |>
  as.data.frame() |>
  pivot_wider(
    names_from = Var2,
    values_from = Freq,
    values_fill = 0
  ) |>
  rename(Species = Var1) |>
  kable(
    caption = "Penguin Count by Species and Island"
  )
```

### Mean Measurements by Species

@fig-species-mean shows mean morphological measurements by species

```{r}
#| label: fig-species-mean
#| fig-asp: 0.618
#| fig-alt: Mean Morphological Measurements by Species

penguins |> 
  filter(!is.na(bill_length_mm), !is.na(bill_depth_mm)) |>
  group_by(species) |>
  summarise(mean_bill_length = mean(bill_length_mm),
            mean_bill_depth = mean(bill_depth_mm),
            mean_flipper_length = mean(flipper_length_mm),
            mean_body_mass = mean(body_mass_g)
            ) |>
  mutate(across(where(is.numeric), \(x) round(x, 1))) |>
  kable(
    caption = "Mean Morphological Measurements by Species",
    align = "r"
  )

```

### Mean Measurements by Sex

@fig-sex-mean shows mean morphological measurements by sex

```{r}
#| label: fig-sex-mean
#| fig-asp: 0.618
#| fig-alt: Mean Morphological Measurements by Sex

penguins |> 
  filter(!is.na(bill_length_mm), !is.na(bill_depth_mm), !is.na(sex)) |>
  group_by(sex) |>
  summarise(mean_bill_length = mean(bill_length_mm),
            mean_bill_depth = mean(bill_depth_mm),
            mean_flipper_length = mean(flipper_length_mm),
            mean_body_mass = mean(body_mass_g)
            ) |>
  mutate(across(where(is.numeric), \(x) round(x, 1))) |>
  kable(
    caption = "Mean Morphological Measurements by Sex",
    align = "r"
  )
```

### Median Measurements by Species

@fig-species-median shows median morphological measurements by species

```{r}
#| label: fig-species-median
#| fig-asp: 0.618
#| fig-alt: Median Morphological Measurements by Species

penguins |> 
  filter(!is.na(bill_length_mm), !is.na(bill_depth_mm), !is.na(sex)) |>
  group_by(species) |>
  summarise(
    median_bill_length = median(bill_length_mm),
    median_bill_depth = median(bill_depth_mm),
    median_flipper_length = median(flipper_length_mm),
    median_body_mass = median(body_mass_g)
  ) |>
  mutate(across(where(is.numeric), \(x) round(x, 1))) |>
  kable(
    caption = "Median Morphological Measurements by Species",
    align = "r"
  )

```

### Median Measurements by Sex

@fig-sex-median shows median morphological measurements by sex

```{r}
#| label: fig-sex-median
#| fig-asp: 0.618
#| fig-alt: Median Morphological Measurements by Sex

penguins |> 
  filter(!is.na(bill_length_mm), !is.na(bill_depth_mm), !is.na(sex)) |>
  group_by(sex) |>
  summarise(
    median_bill_length = median(bill_length_mm),
    median_bill_depth = median(bill_depth_mm),
    median_flipper_length = median(flipper_length_mm),
    median_body_mass = median(body_mass_g)
  ) |>
  mutate(across(where(is.numeric), \(x) round(x, 1))) |>
  kable(
    caption = "Median Morphological Measurements by Sex",
    align = "r"
  )

```

### Mode Measurements by Sex

@fig-sex-mode shows mode morphological measurements by sex

```{r}
#| label: fig-sex-mode
#| fig-asp: 0.618
#| fig-alt: Mode Morphological Measurements by Sex

get_mode <- function(x) {
  ux <- na.omit(unique(x))
  ux[which.max(tabulate(match(x, ux)))]
}

penguins |> 
  filter(!is.na(bill_length_mm), !is.na(bill_depth_mm), !is.na(sex)) |>
  group_by(sex) |>
  summarise(
    mode_bill_length = get_mode(bill_length_mm),
    mode_bill_depth = get_mode(bill_depth_mm),
    mode_flipper_length = get_mode(flipper_length_mm),
    mode_body_mass = get_mode(body_mass_g)
  ) |>
  kable(
    caption = "Mode Morphological Measurements by Sex",
    align = "r"
  )

```

### Mode Measurements by Species

@fig-species-mode shows mode morphological measurements by species

```{r}
#| label: fig-species-mode
#| fig-asp: 0.618
#| fig-alt: Mode Morphological Measurements by Species
#| fig-cap: Mode Morphological Measurements by Species

penguins |> 
  filter(!is.na(bill_length_mm), !is.na(bill_depth_mm), !is.na(sex)) |>
  group_by(species) |>
  summarise(
    mode_bill_length = get_mode(bill_length_mm),
    mode_bill_depth = get_mode(bill_depth_mm),
    mode_flipper_length = get_mode(flipper_length_mm),
    mode_body_mass = get_mode(body_mass_g)
  ) |>
  kable(
    caption = "Mode Morphological Measurements by Species",
    align = "r"
  )
```

### Central Tendency of Penguin Morphological Traits by Species and Sex

@fig-species-summary-all mean, median, and mode morphological measurements by species

```{r}
#| label: fig-species-summary-all
#| fig-asp: 0.618
#| fig-alt: Mean, median, and mode morphological measurements by species
#| fig-cap: Mean, median, and mode morphological measurements by species

get_mode <- function(x) {
  ux <- na.omit(unique(x))
  ux[which.max(tabulate(match(x, ux)))]
}

mean_tbl <- penguins |> 
  filter(!is.na(bill_length_mm), !is.na(bill_depth_mm)) |>
  group_by(species) |>
  summarise(
    `Bill Length (mm)` = mean(bill_length_mm),
    `Bill Depth (mm)` = mean(bill_depth_mm),
    `Flipper Length (mm)` = mean(flipper_length_mm),
    `Body Mass (g)` = mean(body_mass_g),
    .groups = "drop"
  ) |>
  mutate(statistic = "Mean")

median_tbl <- penguins |> 
  filter(!is.na(bill_length_mm), !is.na(bill_depth_mm)) |>
  group_by(species) |>
  summarise(
    `Bill Length (mm)` = median(bill_length_mm),
    `Bill Depth (mm)` = median(bill_depth_mm),
    `Flipper Length (mm)` = median(flipper_length_mm),
    `Body Mass (g)` = median(body_mass_g),
    .groups = "drop"
  ) |>
  mutate(statistic = "Median")

mode_tbl <- penguins |> 
  filter(!is.na(bill_length_mm), !is.na(bill_depth_mm)) |>
  group_by(species) |>
  summarise(
    `Bill Length (mm)` = get_mode(bill_length_mm),
    `Bill Depth (mm)` = get_mode(bill_depth_mm),
    `Flipper Length (mm)` = get_mode(flipper_length_mm),
    `Body Mass (g)` = get_mode(body_mass_g),
    .groups = "drop"
  ) |>
  mutate(statistic = "Mode")

bind_rows(mean_tbl, median_tbl, mode_tbl) |>
  pivot_longer(cols = -c(species, statistic), names_to = "measurement", values_to = "value") |>
  pivot_wider(names_from = statistic, values_from = value) |>
  relocate(measurement, .after = species) |>
  mutate(across(where(is.numeric), \(x) round(x, 1))) |>
  kable(
    caption = "Mean, median, and mode morphological measurements by species",
    align = "r"
  )

```

The mean and median are typically close, indicating relatively symmetrical distributions, but the mode occasionally differs, suggesting some skewness or clustering at certain values (e.g., Gentoo body mass mode = 5550 g vs. mean = 5076 g).

@fig-sex-summary-all mean, median, and mode morphological measurements by sex

```{r}
#| label: fig-sex-summary-all
#| fig-asp: 0.618
#| fig-alt: Mean, median, and mode morphological measurements by sex
#| fig-cap: Mean, median, and mode morphological measurements by sex

get_mode <- function(x) {
  ux <- na.omit(unique(x))
  ux[which.max(tabulate(match(x, ux)))]
}

mean_tbl <- penguins |> 
  filter(!is.na(bill_length_mm), !is.na(bill_depth_mm), !is.na(sex)) |>
  group_by(sex) |>
  summarise(
    `Bill Length (mm)` = mean(bill_length_mm),
    `Bill Depth (mm)` = mean(bill_depth_mm),
    `Flipper Length (mm)` = mean(flipper_length_mm),
    `Body Mass (g)` = mean(body_mass_g),
    .groups = "drop"
  ) |>
  mutate(statistic = "Mean")

median_tbl <- penguins |> 
  filter(!is.na(bill_length_mm), !is.na(bill_depth_mm), !is.na(sex)) |>
  group_by(sex) |>
  summarise(
    `Bill Length (mm)` = median(bill_length_mm),
    `Bill Depth (mm)` = median(bill_depth_mm),
    `Flipper Length (mm)` = median(flipper_length_mm),
    `Body Mass (g)` = median(body_mass_g),
    .groups = "drop"
  ) |>
  mutate(statistic = "Median")

mode_tbl <- penguins |> 
  filter(!is.na(bill_length_mm), !is.na(bill_depth_mm), !is.na(sex)) |>
  group_by(sex) |>
  summarise(
    `Bill Length (mm)` = get_mode(bill_length_mm),
    `Bill Depth (mm)` = get_mode(bill_depth_mm),
    `Flipper Length (mm)` = get_mode(flipper_length_mm),
    `Body Mass (g)` = get_mode(body_mass_g),
    .groups = "drop"
  ) |>
  mutate(statistic = "Mode")

bind_rows(mean_tbl, median_tbl, mode_tbl) |>
  pivot_longer(cols = -c(sex, statistic), names_to = "measurement", values_to = "value") |>
  pivot_wider(names_from = statistic, values_from = value) |>
  relocate(measurement, .after = sex) |>
  mutate(across(where(is.numeric), \(x) round(x, 1))) |>
  kable(
    caption = "Mean, median, and mode morphological measurements by sex",
    align = "r"
  )

```

Male penguins exhibit larger morphological measurements than females across all traits, with the most pronounced differences in body mass (mean = 4545.7 g) and bill length (mean = 45.9 mm). The variation in modal values, such as the lower bill length mode for males (41.1 mm), may indicate skewed or bimodal distributions within sex groups, suggesting underlying variability not captured by mean or median alone.

## Visualization

### Bill length vs Depth

@fig-bill-dims-species below shows the bill length and depth of different penguin species

```{r}
#| label: fig-bill-dims-species
#| fig-asp: 0.618
#| fig-alt: A scatterplot of penguins' bill depth and length, colored by species of penguins. There is a relatively strong, linear association.
#| fig-cap: A scatterplot of penguins' bill depth and length, colored by species of penguins.

ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm )) +
  geom_point(aes(colour = species, shape = species), na.rm = TRUE) +
  labs(
    title = "Bill length and Bill depth of penguin species",
    subtitle = "Adelie have shorter and deeper bill, Gentoo have longer and shallower bill",
    x = "Bill length(mm)",
    y = "Bill depth(mm)",
    colour = "Species",
    shape = "Species"
  ) +
  geom_smooth(method = "lm", formula = y ~ x, na.rm = TRUE) +
  theme_minimal() +
  scale_color_colorblind()
```

The scatterplot shows a clear separation between species: Adelie penguins tend to have shorter, deeper bills; Gentoo have longer, shallower bills; and Chinstrap fall in between. This demonstrates that bill morphology can be a strong species identifier.

### Flipper length vs Body mass

@fig-flipper-bmass-species below shows the flipper length and body mass of different penguin species

```{r}
#| label: fig-flipper-bmass-species
#| fig-asp: 0.618
#| fig-alt: A scatterplot of penguins' flipper length and body mass, colored by species of penguins. There is a relatively strong, linear association.
#| fig-cap: A scatterplot of penguins' flipper length and body mass, colored by species of penguins.

ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g )) +
  geom_point(aes(colour = species, shape = species), na.rm = TRUE) +
  labs(
    title = "Flipper length and Body mass of penguin species",
    subtitle = "Gentoo have longer flipper and are heavier",
    x = "Flipper length(mm)",
    y = "Body mass(g)",
    colour = "Species",
    shape = "Species"
  ) +
  geom_smooth(method = "lm", formula = y ~ x, na.rm = TRUE) +
  theme_minimal() +
  scale_color_colorblind()
```

There is a strong positive correlation between flipper length and body mass. Gentoo penguins generally appear heavier and have longer flippers, while Adelies are lighter with shorter flippers.

### Body mass by species

@fig-bmass-species below shows the body mass distribution by species

```{r}
#| label: fig-bmass-species
#| fig-asp: 0.618
#| fig-alt: A boxplot of penguins' body mass and species.
#| fig-cap: A boxplot of penguins' body mass and species.

ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_boxplot(na.rm = TRUE) +
  labs(
    title = "Body Mass Distribution by Species",
    x = "Species",
    y = "Body Mass (g)"
  ) +
  theme_minimal()
```

Gentoo penguins have the highest body mass, followed by Chinstrap and Adelie. The boxplot highlights a distinct weight difference among the species, with minimal overlap.

### Species by island

@fig-species-island shows species distribution by island

```{r}
#| label: fig-species-island
#| fig-asp: 0.618
#| fig-alt: A barplot of penguins' body mass and species.
#| fig-cap: A barplot of penguins' body mass and species.

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(position = "fill") +
  labs(
    title = "Species Composition by Island",
    x = "Island",
    y = "Proportion",
    fill = "Species"
  ) +
  theme_minimal()
```

Species are distributed unevenly across islands. Adelie penguins are present on all islands, Chinstrap are exclusive to Dream, and Gentoo are only found on Biscoe.

### Flipper length vs body mass by island

@fig-flipper-bmass-species-island below shows a scatterplot of flipper length by body mass faceted by island

```{r}
#| label: fig-flipper-bmass-species-island
#| fig-asp: 0.618
#| fig-alt: A scatterplot of flipper length by body mass faceted by island.
#| fig-cap: A scatterplot of flipper length by body mass faceted by island.

ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species, shape = species), na.rm = TRUE) +
  facet_wrap(~island) +
    labs(
    title = "Flipper Length vs. Body Mass by Island",
    x = "Flipper Length (mm)",
    y = "Body Mass (g)",
    color = "Species",
    shape = "Species"
  ) +
  scale_color_colorblind()
```

The relationship between flipper length and body mass holds consistently across islands, with Biscoe showing higher body mass and flipper length largely influenced by the Gentoo population.

### Bill length and depth by sex

@fig-billdims-sex-island shows a scatterplot of bill length and depth faceted by sex.

```{r}
#| label: fig-billdims-sex-island
#| fig-asp: 0.618
#| fig-alt: A scatterplot of bill length and depth faceted by sex.
#| fig-cap: A scatterplot of bill length and depth faceted by sex.

ggplot(filter(penguins, !is.na(sex)), aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point(aes(color = sex), na.rm = TRUE) +
  facet_wrap(~species) +
  labs(
    title = "Bill Length vs. Bill Depth by Species and Sex",
    subtitle = "Color shows sex; panels represent species.",
    x = "Bill Length (mm)",
    y = "Bill Depth (mm)",
    color = "Sex"
  ) +
  theme_minimal() +
  scale_color_colorblind()
```

Across both sexes, males generally exhibit longer and deeper bills compared to females. The distribution patterns are similar, but males show more variability and larger measurements.

### Body mass by species and sex

@fig-bmass-species-sex shows a boxplot of species by body mass colored by sex.

```{r}
#| label: fig-bmass-species-sex
#| fig-asp: 0.618
#| fig-alt: A boxplot of species by body mass colored by sex.
#| fig-cap: A boxplot of species by body mass colored by sex.

ggplot(filter(penguins, !is.na(sex)), aes(x = species, y = body_mass_g, fill = sex)) +
  geom_boxplot(na.rm = TRUE, position = position_dodge(width = 0.75)) +
  labs(
    title = "Body Mass by Species and Sex",
    subtitle = "Boxplots grouped by species and filled by sex",
    x = "Species",
    y = "Body Mass (g)",
    fill = "Sex"
  ) +
  theme_minimal() +
  scale_fill_colorblind()
```

Within each species, males tend to be heavier than females. Gentoo males are the heaviest group overall, followed by Chinstrap and Adelie males.
